TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-09 src/paper.tex
update

The discretization approach for Target Speech Extraction (TSE) has received limited attention in the literature. One of the earliest works to introduce discrete tokens in the context of TSE is presented by \cite{gen_tse}, which investigates the use of self-supervised learning (SSL) models such as HuBERT and vq-wav2vec \cite{}. However, this study does not explore the WavLM SSL model, which has shown superior performance in speech separation tasks. Additionally, it primarily utilizes the output from a single layer of the SSL model for discretization. Furthermore, the evaluation in \cite{gen_tse} focuses solely on speech quality, neglecting key metrics such as speech intelligibility and speaker similarity, both of which are critical for the TSE task. To address these limitations, we propose TSELM, a novel approach that leverages multiple hidden layers of WavLM for discretization and incorporates evaluation metrics for speech intelligibility and speaker similarity.

@@

It is shown that metrics like PESQ, SI-SNR, STOI do not reflect speech quality of output of 
vocoders due to the fact that the vocoder output does not focus strictly on frame alignment
\cite{tokensplit,selm}. We used DNSMOS \cite{dnsmos} to measure the speech quality, and differential 
word error 
rate (dWER) \cite{dwer} to measure the speech intelligibility. For speaker similarity, we use the 
public WeSpeaker \cite{wespeaker}.

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech. It serves as the upper bond for our model.  Besides TSELM, we conduct three main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoConcat using the same training data. 
For Continuous-WavLM-6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross attention  
and LM without discretization. 
The concatenation strategy is stilled applied to the mixed speech. Mean Square Error (MSE) loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. We call 
it hybrid because the mixture speech retains continuous features. In TSELM-S-NoConcat, we utilize the TSELM-S model architecture but without concatenation strategy.

@@

We observe a significant increase in dWER when the mixture audio is not concatenated with the reference, as seen in the TSELM-S-NoConcat results in Table \ref{main_exp}. For WavLM to effectively perform target speaker separation, the input audio must follow specific conditions: the mixture should be less than 50\% of the total length, and the first utterance should be the target speaker. Under these conditions, WavLM outputs a slightly denoised embedding that emphasizes the target speaker. When the entire input is a mixture, however, we found that WavLM sometimes extracts the wrong speaker.
Our current concatenation strategy is inspired by SELM's \cite{selm} success in speech denoising and aims to transform the target speaker extraction task into a more difficult speech enhancement problem leveraging WavLM's denoising capability.

>>>>

The discretization approach for Target Speech Extraction (TSE) has received limited attention in the literature. One of the earliest works to introduce discrete tokens in the context of TSE is presented by \cite{gen_tse}, which investigates the use of self-supervised learning (SSL) models such as HuBERT and vq-wav2vec \cite{vq_wav2vec}. However, this study does not explore the WavLM SSL model, which has shown superior performance in speech separation tasks \cite{wavlm}. Additionally, it primarily utilizes the output from a single layer of the SSL model for discretization. Furthermore, the evaluation in \cite{gen_tse} focuses solely on speech quality, neglecting key metrics such as speech intelligibility and speaker similarity, both of which are critical for the TSE task. To address these limitations, we propose TSELM, a novel approach that leverages multiple hidden layers of WavLM for discretization and incorporates evaluation metrics for speech intelligibility and speaker similarity.

@@

It is shown that metrics like PESQ, SI-SNR, STOI do not reflect speech quality of output of 
vocoders due to the fact that the vocoder output does not focus strictly on frame alignment
\cite{tokensplit,selm}. We use DNSMOS \cite{dnsmos} to measure the speech quality, and differential 
word error 
rate (dWER) \cite{dwer} to measure the speech intelligibility. For speaker similarity, we use the 
public WeSpeaker \cite{wespeaker}.

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech. It serves as the upper bond for our model.  Besides TSELM, we conduct three main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoCat using the same training data. 
For Continuous-WavLM-6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross attention  
and LM without discretization. 
The concatenation strategy is stilled applied to the mixed speech. Mean Square Error (MSE) loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. We call 
it hybrid because the mixture speech retains continuous features. In TSELM-S-NoCat, we utilize the TSELM-S model architecture but without concatenation strategy.

@@

We observe a significant increase in dWER when the mixture audio is not concatenated with the reference, as seen in the TSELM-S-NoCat results in Table \ref{main_exp}. For WavLM to effectively perform target speaker separation, the input audio must follow specific conditions: the mixture should be less than 50\% of the total length, and the first utterance should be the target speaker. Under these conditions, WavLM outputs a slightly denoised embedding that emphasizes the target speaker. When the entire input is a mixture, however, we found that WavLM sometimes extracts the wrong speaker.
Our current concatenation strategy is inspired by SELM's \cite{selm} success in speech denoising and aims to transform the target speaker extraction task into a more difficult speech enhancement problem leveraging WavLM's denoising capability.
